---
title: Intro to Categorical Analysis in R
author: R package build
date: '2021-10-20'
slug: []
categories:
  - R
tags: []
subtitle: ''
summary: 'Categorical Data Analysis in SAS....in R'
authors: []
lastmod: '2021-10-20T19:24:09-04:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
bibliography: ['refs.bib']
---

Load the req packages
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(gmodels)
```
This is a first post of (hopefully) many that works through some of 'Categorical Data Analysis in SAS' 3rd ed. by Stokes et al.[@stokes_categorical_2012] Since that book is obviously written in SAS for SAS, there is no R code. So for people interested in using R to perform the same tests that book covers, there is this post. I hope it is useful.

First make data. We are working to emulate loading summary data. In SAS it would load in like this:
```{r, eval=FALSE}
data exdata;
input center $ treat $ outcome $ count @@;
datalines;
1 new fav 64 1 new unfav 36
1 std fav 53 1 std unfav 47
2 new fav 35 2 new unfav 15
2 std fav 22 2 std unfav 28
;
run;
```

but in R we need to make data like this:
```{r}
exdata<-data.frame(center=c(1,1,1,1,2,2,2,2),
                 treat=c("new","new","std","std",
                         "new","new","std","std"),
                 outcome=c("fav","unfav","fav","unfav",
                           "fav","unfav","fav","unfav"),
                 count=c(64,36,53,47,35,15,22,28))

```
 
And the data frame looks like this:

```{r}
print(exdata)
```
However, as far as I know there is no easy way to use this moving forward. In `SAS` you would use a `weight` statement that would let `SAS` know you want to weight the data, but `R` doesnt have a similar function that I can tell, so I made a rudimentary one. This  function makes a dataframe that has the correct number of observations for each combination of variables so we dont need to rely on a weight statement.

First a function to load in the data.frame. This will skirt the need to use `weight` in `PROC`s in SAS.
```{r}

sassy_summary_data <- function(data,#data to use should be a dataframe
                               var1,#variables other than number of counts to be considered
                               var2,
                               var3,
                               var4,
                               var5,
                               var_num,#number of variables
                               weight){
    df<-vector()
    print(data)
    #testpre<-dplyr::count(data,data[,1],
                          #data[,2],
                          #data[,3],
                          #wt=data[,4])
    reps<-2^var_num
    library(tidyverse)
    for (i in 1:reps){
        test<-data[i,] %>% slice(rep(1:n(),
                                        each=data$count[i]))
        df<-rbind(df,test)
    }
    return(df)
}
```

Then if we call that function:
```{r}
newexdata<-sassy_summary_data(data=exdata,
                              "outcome",
                              "treat",
                              "center",
                              weight="count",
                              var_num=3)#three vars outcome, treat, and center
```

We can see we have a new data frame with the same summary numbers for each combination of variables, but now it can be used by R to do some categorical analysis.

```{r}
newexdata %>% group_by_all() %>%  count()
```

There are other ways to load the data too, that dont require my work around above. For instance you can load the load as reported first and then use `xtabs` and `ftable` to make an analyzable table too.

Load data again
```{r}
exdata2<-data.frame(center=c(1,1,1,1,2,2,2,2),
                 treat=c("new","new","std","std",
                         "new","new","std","std"),
                 outcome=c("fav","unfav","fav","unfav",
                           "fav","unfav","fav","unfav"),
                 count=c(64,36,53,47,35,15,22,28))
```


use xtabs to make crosstabulation
```{r}
xtabs(data=exdata2, count~center +treat+outcome)
```

and use ftable to flatten it
```{r}
flat_table<-ftable(xtabs(data=exdata2, count~center +treat+outcome))
print(flat_table)
```

the above would make a table similar to `SAS` in `PROC FREQ` if you had the following statment: `tables center*treat*outcome`

This method, to me, seems a little more of a pain since you need to remake the tables for different analysis


Now Lets to some analysis. The second chapter is 2x2 tables.

## Chp2: 2x2 Tables

Mantel-Haenszel chi-square statistic test


## In SAS

```{r eval=FALSE}
PROC FREQ data=exdata order=data;
weight count;
tables center*treat*outcome/cmh chisq; 
#for marginal without accounting for center, remove center from 
#tables statement
run;
```


For reference the SAS output, below, shows for non corrected version:

Marginal (ie not adjusted for center):

```{r}

 Statistic                     DF       Value      Prob
     ƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒ
     Mantel-Haenszel Chi-Square     1      7.8555    0.0051
```


And Adjusted for Center:

```{r eval=FALSE}
   Cochran-Mantel-Haenszel Statistics (Based on Table Scores)
 
 Statistic    Alternative Hypothesis    DF       Value      Prob
 ƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒƒ
     1        Nonzero Correlation        1      7.8307    0.0051
     2        Row Mean Scores Differ     1      7.8307    0.0051
     3        General Association        1      7.8307    0.0051
```


## in R 

marginal:

```{r}
table(newexdata$outcome, newexdata$treat) %>%  DescTools::MHChisqTest()
```
and adjusting for center:

```{r}
#can pass it a 3-dimensional array
table(newexdata$treat,
      newexdata$outcome,
      newexdata$center) %>%#center listed last because it is the strata
  mantelhaen.test(correct=FALSE) 
```

```{r}
#or can pass it factors
mantelhaen.test(newexdata$treat,
                newexdata$outcome,
                newexdata$center, correct=TRUE)
```

And we see we can do this easily in R

```{r}
sI<-sessionInfo()

sI$R.version$version.string
sI$R.version$platform
sI$basePkgs#packages used
names(sI$otherPkgs) #packages used
```

# References


