---
title: "How to Visualize Binary Outcomes in Research"
author: "TF"
date: '2018-06-16'
output:
  html_document:
    df_print: paged
header:
  caption: ''
  image: ''
slug: how-to-visualize-binary-outcomes-in-research
tags:
- R
- Data analysis
- dplyr
categories: []
---

In my research I frequently want visualize outcomes over groups. The nature of clinical reserach means that often times outcomes are binary and I want to find the proportion of binary outcomes over groups. This was not straight forward for me, so I decided to make a post about it.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

###load libraries
```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(ggpubr)
```

###create the variables
```{r}
set.seed(15762)
death<-rbinom(1000,1,0.03)
readmit <-rbinom(1000,1, 0.1)
id<-rbinom(1000,5,0.2)
```

###combine into dataframe
```{r}
x <- data.frame(death,readmit,id)
names(x) <- c("death", "readmit", "id")
```
###check the results
```{r}
head(x)
```
Now here is where it was tricky for me. If you, like I typically do, work with large data then you will need to extract the components you want. Typically for the specific outcome I am interested in, i'll use `na.omit(df)` to restrict to only observations with all outcomes observed; excluding NAs

So once you do that you want to select out the binary outcome you are interested in, and I will be using death as the outcome to evaluate. 

###pull out the observations I am interested in the proportion of death across different IDs
```{r}
death_x <-dplyr::select(x, death, id)
```
###create proportion table to get proportion of deaths across different IDs
```{r}
death_prop <- death_x %>% count(death, id) %>% mutate(prop=prop.table(n))
```

###now just look at the proportion of those that died; this next step basically makes sure we only visualize the proportion of death
```{r}
death_prop <- dplyr::filter(death_prop, death==1)
```

###then just graph it; ive chosen to use a lollipop plot, because I think that is the best
```{r}
ggplot(death_prop, aes(x=id, y=prop)) +
  geom_point() + 
  geom_segment( aes(x=id, xend=id, y=0, yend=prop))+theme_classic()+geom_hline(aes(yintercept=0))+labs(x="ID #", y="Proportion of Death", title="Variability in Death over ID")
```
